---
title: "4 Flexible parametric survival"
format: 
  html:
    embed-resources: true
    code-overflow: wrap
toc: true
toc-depth: 3
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

## What is a flexible parametric survival model?

While standard parametric models are useful, they are often limited by rigid assumptions about the shape of the hazard function. Flexible parametric models, famously introduced by **Patrick Royston and Catherine Parmar in 2002**, were developed to bridge the gap between these standard models (like the Weibull) and semi-parametric approaches like Cox proportional hazards model. As parametric model, they provide a fully defined mathematical formula for the entire survival curve, allowing for easy extrapolation and absolute risk measures. As a Cox model, they are flexible enough to fit almost any data shape.

The flexibility in these models comes from the use of **restricted cubic splines** (polynomials) to model the log-cumulative hazard or log-hazard function rather than the hazard itself. Rather than assuming a simple linear trend, the model uses these splines joined at specific **'knots'** to capture complex, irregular, or multi-modal hazard shapes that simpler models would miss.

Unlike piecewise models, which require splitting data into arbitrary time intervals, flexible parametric models provide smooth, continuous estimates of both survival and hazard functions.

While 'flexible parametric model' is the broad category for any parametric model that uses splines, the Royston-Parmar (RP) model is the most famous implementation of this idea.

------------------------------------------------------------------------

## Why use flexible parametric survival models?

**Key advantages of flexible parametric models**

-   **Wider hazard variety:** Beyond the other benefits of parametric models, flexible parametric models can capture a wider range of hazard shapes (and thus biological realities) than standard parametric models
-   **Time-dependent effects:** Often the standard choice when **proportional hazards fail**, as they easily allow for time-dependent effects (e.g., the impact of a treatment changes over the duration of the study)
-   **Absolute measures and extrapolation:** Because they are fully parametric, they allow for extrapolation and calculation of **absolute measures of effect** (like absolute risks and rates or life expectancy) which are difficult to obtain from Cox models.

Flexible parametric models are thus a good methodological choice when:

**1. The hazard function is complex and of interest:** When $h(t)$ is non-monotonic, simpler models may fail to capture the nuance

**2. A single hazard ratio fails to capture the complete clinical story:**: The HR is a summary measure; it doesn't tell how risk evolves over time

**3. The proportional hazars assumption is violated or inappropriate:** Flexible parametric models allow for non-proportional hazards and time-dependent effects

------------------------------------------------------------------------

## Key concepts in flexible parametric models

**Polynomials**

The foundation of flexibility starts with high-order functions. **Polynomials** can be used to model a **non-linear function**. A **cubic polynomial** can model a non-linear function using the form:

$$y_i = \beta_0 + \beta_1x_i + \beta_2x_i^2 + \beta_3x_i^3$$

**Splines**

When a single polynomial is too rigid to fit the entire range of data, we use **splines**. They are a series of polynomials, essentially **piecewise polynomials**. **Cubic splines are piecewise cubic polynomials**. They are thus flexible mathematical functions composed of several polynomial segments joined together at specific time points called **knots**. Each segment models a specific time interval, allowing for local flexibility.

**Restricted cubic splines (RCS)**

Restricted cubic splines, also known as **natural splines**, are the "gold standard" for flexible parametric modelling because they add necessary constraints:

-   **Continuity:** They are defined to be continuous at the knots.
-   **Smoothness:** They have continuous first and second derivatives at the knots (i.e., no sharp 'corners').
-   **Linear Constraints:** They are constrained to be linear beyond the *boundary knots*.
-   **Boundary Placement:** Boundary knots are typically placed at the minimum and maximum of the (uncensored) event times to prevent erratic behavior at the ends of the data.

By forcing the spline function to be linear at the tails (before the first knot and after the last knot), we avoid the common issue where polynomials 'fly off' to infinity in areas where we have very little data.

**Knot placement and number of knots**

The flexibility of the model is determined by the placement and number of knots. *Internal Knots* are usually placed at equally spaced centiles of the distribution of event times. For example, if using one internal knot, it is typically placed at the median survival time. Choosing the number of knots (denoted as $k$) involves a trade-off between capturing detail and avoiding overfitting.

------------------------------------------------------------------------

## Spline function and basis calculation

The flexibility of a restricted cubic spline is derived from its **basis functions**. Let $x$ be a function of time — typically $x = \log(t)$.

**Notation and knots**

-   **Boundary knots:** $k_{min}$ and $k_{max}$ represent the minimum and maximum of the log event times.
-   **Internal knots:** $k_1, \cdots, k_m$ represent the location of $m \ge 0$ internal knots, with $k_j$ representing the location of the $j^{th}$ internal knot.

**The spline function**

The spline function $s(x,\gamma)$ is defined as:

$$s(x,\gamma) = \gamma_0 + \gamma_1 x + \sum_{j=1}^{m} \gamma_{j+1} v_j(x)$$

**Calculating the basis functions**

For each internal knot $k_j$, the $j^{th}$ **basis function** $v_j(x)$ is calculated using a combination of cubic terms:

$$v_j(x) = (x - k_j)^3_+ - \lambda_j(x - k_{min})^3_+ + (\lambda_j - 1)(x - k_{max})^3_+$$ *Key components:*

-   **The truncated power function:** The term $(x - k)_+^3$ equals $(x - k)^3$ if $x > k$, and $0$ if $x \leq k$.
-   **The weighting factor (**$\lambda_j$): This is calculated based on the distance between the knots:

$$\lambda_j = \frac{k_{max} - k_j}{k_{max} - k_{min}}$$

::: callout-important
**Why the weighting factor matters**

The $\lambda_j$ factor is what makes the cubic spline **restricted**. It is mathematically engineered to force the cubic ($x^3$) and quadratic ($x^2$) components of the function to sum to zero beyond the boundary knots. This ensures that the model remains **linear** at the tails, preventing unstable predictions where data is sparse.
:::

------------------------------------------------------------------------

## How do flexible parametric models work?

**Integration into regression**

Regression splines are highly versatile because they can be incorporated into any regression model that utilises a **linear predictor**. This allows us to move beyond rigid assumptions while maintaining a familiar framework.

**The scale of modelling**

While splines can be used to model either the log-hazard or the log-cumulative hazard function, most flexible parametric models (such as the Royston-Parmar model) focus on the **log-cumulative hazard**:

-   **Monotonicity:** It is mathematically easier for a cubic spline to fit a smooth, monotonically increasing curve like the cumulative hazard.
-   **Stability:** Estimating the cumulative hazard often leads to more stable numerical estimation compared to the instantaneous hazard.

**The importance of the log-time transformation**

In these models, it is generally superior to use $\log(t)$ rather than raw time ($t$). This transformation serves two critical purposes:

-   **1. Distributional balancing:** As survival times are often positively skewed, the log transformation 'compresses' the long right tail of survival distributions (where events are sparse) and 'stretches' the early period (where many events typically occur).
-   **2. Improved Extrapolation:** A linear trend on the log-time scale is more biologically and statistically plausible for extrapolation beyond the boundary knots compared to a linear trend on the raw time scale.

::: callout-note
### Why using $\log(t)$ is important?

Using $\log(t)$ ensures that the 'restricted' part of our cubic spline (the linear tail) behaves predictably. Extrapolating a linear trend in log-time is equivalent to a Weibull-like tail, which is a standard and safe assumption in survival analysis.
:::

------------------------------------------------------------------------

## Review: The Weibull model

Royston and Parmar designed their flexible parametric model as a direct extension of the Weibull model. They used the Weibull model because of its relationship with the log-cumulative hazard that is linear with log-time.

Before diving into Royston-Parmar model, let's recall the framework of the Weibull model:

-   **Survival function:** $$S(t) = \exp(-\lambda t^\gamma)$$

-   **Log cumulative hazard:** $$\log(-\log(S(t))) = \log(H(t)) = \log(\lambda) + \gamma \log(t)$$ $\rightarrow$ This is a **linear function** of $\log(t)$.

-   **Introducing covariates gives:** $$\log(H(t|x_i)) = \log(\lambda) + \gamma \log(t) + x_i \beta$$

------------------------------------------------------------------------

## The Log-cumulative hazard framework in flexible parametric models

The Royston-Parmar model starts with the linear relationship from the Weibull model:

$$\log(H(t)) = \gamma_0 + \gamma_1 \log(t) + x_i \beta$$ It then adds flexibility to it where needed. That is, rather than assuming a strictly linear relationship with $\log(t)$, it uses **restricted cubic splines (RCS)** for $\log(t)$:

$$\log(H(t)) = \gamma_0 + \gamma_1 \log(t) + \sum_{j=1}^{k} \gamma_{j+1} v_j(log(t)) + x_i \beta$$

Using the Weibull model as a starting point, the flexible parametric models use cubic splines to 'bend' that line into more complex shapes. This allows the model to capture deviations from the Weibull assumption (i.e., when the hazard isn't strictly monotonic).

The complexity of the model is governed by the number of knots ($k$). Adding more knots allows for more 'bends' in the log-cumulative hazard curve.

------------------------------------------------------------------------

## Generalisations of flexible parametric models

More generally, in flexible parametric models, a **link function** $g(\cdot)$ transforms the survival function so it can be modelled as a **restricted cubic spline function of log-time**:

$$g(S(t|x)) = s(\log(t), \gamma) + x\beta$$ The link function defines the scale on which the spline and covariate effects are additive. The following link functions appear in the Royston-Parmar framework:

-   **Complementary log-log:** Typically, we use $g(S(t|x)) = \log(-\log(S(t|x))) = \log(H(t|x))$, the **log-cumulative hazard** which gives a **proportional hazards model**.
-   **Logit:** Using $g(S(t|x_i)) = \log(S(t|x)^{-1} - 1)$, **log-cumulative odds**, gives a **proportional odds model**.
-   **Probit:** Using $g(S(t|x_i)) = \Phi^{-1}(S(t|x))$, the **inverse normal cumulative distribution function (CDF)**, gives a **normal/probit model**.

The normal/probit model (often referred to as the log-normal flexible parametric model) shifts the focus from the hazard rate to the probability of the event occurring by a certain time.

If we assume linearity with log-time (0 internal knots), we are back to a standard parametric model, i.e., Weibull model, log-logistic model, and log-normal model covered before.

Adding internal knots allows the cubic splines to 'bend' the log-time relationship, accommodating hazards that do not follow these strict parametric shapes.

::: callout-important
The choice of link function determines the underlying framework and the 'baseline' distribution the model reverts to if **0 internal knots** are used:

| Link Function $g(S(t))$ | Model Type           | Reverts to (0 knots) |
|:------------------------|:---------------------|:---------------------|
| $\log(-\log(S(t)))$     | Proportional Hazards | **Weibull**          |
| $\log(S(t)^{-1} - 1)$   | Proportional Odds    | **Log-logistic**     |
| $\Phi^{-1}(1 - S(t))$   | Normal/Probit        | **Log-normal**       |
:::

------------------------------------------------------------------------

## Modelling time-dependent effects

Flexible parametric models include the effect of time as covariates in the linear predictor, so time-dependent effects can be included by fitting interactions between the covariate of interest and the covariates defining the effect of time:

$$g(S(t|x)) = s(\log(t), \gamma) + x\beta + s(\log(t), \delta) \cdot x$$

For any time-dependent effect there is thus an interaction between the covariate and the spline variables.

This allows the effect of covariates to vary over time, which is a universal feature of flexible parametric models that allows us to move beyond restrictive assumptions:

-   **In PH models:** Relaxes the proportional hazards assumption (gives Non-PH).
-   **In PO models:** Relaxes the proportional odds assumption (gives Non-PO).
-   **In Probit/AFT models:** Allows the acceleration factor to vary over time.

It is important to think about why the effect might be time-dependent. Is it due to true causal effect or perhaps just due to unobserved frailty.

## Data example: Fitting a flexible parametric model

In practice, R packages like [flexsurv](https://cran.r-project.org/web/packages/flexsurv/vignettes/flexsurv.pdf) or [rstpm2](https://www.rdocumentation.org/packages/rstpm2/versions/1.7.1) handle the estimation process.

**Flexible parametric proportional hazards model**

We can fit flexible parametric proportional hazards model using [flexsurvspline](https://www.rdocumentation.org/packages/flexsurv/versions/2.3.2/topics/flexsurvspline) function from the [flexsurv](https://cran.r-project.org/web/packages/flexsurv/vignettes/flexsurv.pdf) package, by specifying `scale = hazard`.

First, fitting flexible parametric proportional hazards model with 0 knots ($k = 0$) shows that it reverts back to Weibull model:

```{r,warning=FALSE}

# make sure you have installed the MASS and flexsurv packages
# install.packages("MASS")
# install.packages("flexsurv")
# load the flexsurv package
# note that when you load the flexsurv package, it also automatically loads the 
# survival package, as it is built directly on top of the survival package's 
# infrastructure, and needs the Surv() function and the underlying survival object 
# logic to function
library(flexsurv) # flexsurvreg for parametric models and flexsurvspline for
# flexible parametric models
library(MASS) # for gehan data

# read in the gehan data for the leukaemia trial in MASS package gehan
leukdata <- MASS::gehan

# Setting control group as the reference group
leukdata$treat <- factor(leukdata$treat, levels = c("control", "6-MP"))

# WEIBULL PH MODEL

# 1. Fitting the Weibull model using flexsurv package
surv_wei1 <- flexsurvreg( Surv(time, cens) ~ treat, data = leukdata, dist = "weibullPH")
print(surv_wei1)

# 2. Fitting the flexible parametric proportional hazards model with 0 internal knots
# k = 0 means only boundary knots exist, making it a linear model on the log-log scale
surv_spline_PH_0 <- flexsurvspline(Surv(time, cens) ~ treat, 
                                data = leukdata, 
                                k = 0, 
                                scale = "hazard")
print(surv_spline_PH_0)

# 3. Compare log-Likelihoods
print(surv_wei1$loglik)
print(surv_spline_PH_0$loglik)

```

You can see that their log-likelihoods are the same and the parameters are the same except for the `scale` parameter that is reported on a different scale (log(0.0464) = -3.0707).

`Flexsurvspline` output can also be converted into the Weibull AFT parameters in the following way:

```{r,warning=FALSE}

convert_spline_to_weibull <- function(spline_model) {
  # Extract gamma coefficients
  # gamma0 is the intercept, gamma1 is the slope of log(t)
  g0 <- surv_spline_PH_0$res["gamma0", "est"]
  g1 <- surv_spline_PH_0$res["gamma1", "est"]
  
  # Transformation logic
  weibull_shape <- g1
  weibull_scale <- exp(-g0 / g1)
  
  return(data.frame(
    Parameter = c("Weibull Shape (p)", "Weibull Scale (lambda)"),
    Value = c(weibull_shape, weibull_scale)
  ))
}

my_weibull_params <- convert_spline_to_weibull(fit_spline_0k)
print(my_weibull_params)

```

Similarly, we can fit flexible parametric proportional hazards model with 1, 2, or 3 knots, varying the $k$:

```{r,warning=FALSE}

# Fitting the flexible parametric proportional hazards model with 
# 1 internal knot
surv_spline_PH_1 <- flexsurvspline(Surv(time, cens) ~ treat, 
                                data = leukdata, 
                                k = 1, 
                                scale = "hazard")
print(surv_spline_PH_1)

# 2 internal knots
surv_spline_PH_2 <- flexsurvspline(Surv(time, cens) ~ treat, 
                                data = leukdata, 
                                k = 2, 
                                scale = "hazard")
print(surv_spline_PH_2)

# 3 internal knots
surv_spline_PH_3 <- flexsurvspline(Surv(time, cens) ~ treat, 
                                data = leukdata, 
                                k = 3, 
                                scale = "hazard")
print(surv_spline_PH_3)

```

## How many knots to use?

The flexible parametric models are usually not nested and can be compared using the AIC and BIC:

-   If the improvement in fit (likelihood) is greater than the penalty for the new parameter, the AIC/BIC decreases (model with more knots is better)
-   If the improvement in fit is marginal and does not justify the extra parameter, t he AIC/BIC increases (model with fewer knots is better)

From the above outputs, we can see that the AIC increases with each model and additional knot, hence supporting standard Weibull model as the optimal model fit in these data.

Note that as the Weibull model is nested within the flexible parametric proportional hazards model, we can also use the likelihood ratio test to determine which model provides a better fit for the data:

```{r,warning=FALSE}

# Make sure you have installed the lmtest package
# install.packages("lmtest")
# Load the lmtest package
library(lmtest)

# Likelihood Ratio Test

# It is recommended to list the most complex model first
lrtest(surv_spline_PH_1, surv_wei1)

```

Note also that the default knot positions have been shown to work fairly well.
